Supporting Information:\\Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes

### 2021-05-07 manuscriptPGPEsuppl.tex

\title{ Supporting Information:\\Conformational plasticity of phospholipid headgroups in simulations and experiments}

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Code for all atom simulations is available in Ref.~\citenum{MATCHgit} (\texttt{scripts/calcOrderParameters.py}).
For united atom simulations, we first constructed trajectories including hydrogens with ideal geometry using either \texttt{buildH} program~\cite{buildH} or (\texttt{scratch/opAAUA\_prod.py}) in  Ref.~\citenum{MATCHgit}, and the order parameters were then calculated from these trajectories. This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildHvalidation,piggot17}.\\
%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\

@@

\section{Author contributions}
\noindent
{\it Am{\'e}lie Bacle} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the Berger Force Field.
%% Please write description of your contribution here

\noindent
{\it Pavel Buslaev}
%% Please write description of your contribution here
performed the analysis of dihedrals and isomers of lipids. Analysed lipid structures from Protein Data Bank. Prepared panels for figures 2 and 4. Participated in discussions.

\noindent
{\it Rebeca Garc{\'i}a Fandi{\~n}o}
designed and supervised the molecular dynamics simulations carried out by Antonio Pe{\'o}n, and contributed to some discussions.
%% Please write description of your contribution here

@@

\noindent
{\it Tiago M. Ferreira}
was responsible for the solid-state NMR experiments/figures and took part in writing the manuscript.

\noindent
{\it Patrick F.J. Fuchs} supervised Paula Mil{\'a}n Rodr{\'i}guez, Am{\'e}lie Bacle and Chris Papadopoulos, created the buildH software, contributed to many discussions.
%% Please write description of your contribution here

\noindent
{\it Ivan Gushchin} supervised Pavel Buslaev and contributed to the analysis of lipid structures in the Protein Data Bank.
%% Please write description of your contribution here

\noindent
{\it Matti Javanainen} set up and performed simulations using CHARMM36, Slipids, and MacRog lipid models. He contributed to the organization of the manuscript.
%% Please write description of your contribution here

\noindent
{\it Anne M. Kiirikki} set up and performed POPC:POPG(1:1) and POPC:POPG(4:1) simulations with CHARMM36 and POPC:POPG(4:1) with Lipid17ecc. She also contributed to the NMRlipids databank.
%% Please write description of your contribution here

@@

\noindent
{\it Josef Melcr}
created the analysis script \texttt{scripts/calcOrderParameters.py}, 
prepared the Lipid17ecc topology, 
and contributed to discussions and manuscript preparation.

\noindent
{\it Paula Mil{\'a}n Rodr{\'i}guez} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the MacRog Force Field. She also fixed the initial force field (itp) files prior to the simulations.
%% Please write description of your contribution here

\noindent
{\it Markus S. Miettinen}
%% Please write description of your contribution here

@@

\noindent
{\it {\'A}ngel Pi{\~n}eiro}
Created the opAAUA\_prod.py code for the calculation of order parameters. Contributed to some discussions. 
%% Please write description of your contribution here

\noindent
{\it Salla I. Virtanen}
%% Please write description of your contribution here

>>>>

\title{ Supporting Information:\\Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes}

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Code for all atom simulations is available in Ref.~\citenum{MATCHgit} (\texttt{scripts/calcOrderParameters.py}).
For united atom simulations, we first constructed trajectories including hydrogens with ideal geometry using either \texttt{buildH} program~\cite{buildH} or (\texttt{scratch/opAAUA\_prod.py}) in  Ref.~\citenum{MATCHgit}, and the order parameters were then calculated from these trajectories. This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildHvalidation,piggot17}.\\

@@

\section{Author contributions}
\noindent
{\it Am{\'e}lie Bacle} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the Berger force field.
%% Please write description of your contribution here

\noindent
{\it Pavel Buslaev}
%% Please write description of your contribution here
contributed to the analysis of dihedrals and isomers of lipids from MD simulations, and to the analysis of lipid structures from the Protein Data Bank. Prepared panels for figures 2 and 4.

\noindent
{\it Rebeca Garc{\'i}a Fandi{\~n}o}
designed and supervised the molecular dynamics simulations carried out by Antonio Pe{\'o}n.
%% Please write description of your contribution here

@@

\noindent
{\it Tiago M. Ferreira}
was responsible for the solid-state NMR experiments. Prepared figures~\ref{HGorderParameters},~\ref{POPEspectra}, and~\ref{POPGspectra}, and took part in writing the manuscript.

\noindent
{\it Patrick F.J. Fuchs} supervised Paula Mil{\'a}n Rodr{\'i}guez, Am{\'e}lie Bacle and Chris Papadopoulos, created the buildH software and contributed to the calculation of lipid isomers. Provided comments on the manuscript.
%% Please write description of your contribution here

\noindent
{\it Ivan Gushchin} supervised Pavel Buslaev and contributed to the analysis of lipid structures in the Protein Data Bank. Provided comments on the manuscript.
%% Please write description of your contribution here

\noindent
{\it Matti Javanainen} set up and performed simulations using CHARMM36, Slipids, and MacRog lipid models. Provided comments on the manuscript.
%% Please write description of your contribution here

\noindent
{\it Anne M. Kiirikki} developed the preliminary version of the NMRlipids databank, which was used in several analyses. Set up and performed POPC:POPG(1:1) and POPC:POPG(4:1) simulations with CHARMM36 and POPC:POPG(4:1) with Lipid17ecc.
%% Please write description of your contribution here

@@

\noindent
{\it Josef Melcr}
%created the analysis script \texttt{scripts/calcOrderParameters.py}, 
contributed to the preparation of the Lipid17ecc parameters. 
Provided comments on the manuscript.

\noindent
{\it Paula Mil{\'a}n Rodr{\'i}guez} corrected errors in available MacRog force field parameters. Set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the MacRog force field.
%% Please write description of your contribution here

\noindent
{\it Markus S. Miettinen}
made figures~\ref{HGorderParametersPE} and~\ref{HGorderParametersPOPG}. Provided comments on the manuscript.
%% Please write description of your contribution here

@@

\noindent
{\it {\'A}ngel Pi{\~n}eiro}
created the opAAUA\_prod.py code for the calculation of order parameters from united atom trajectories.
%% Please write description of your contribution here

\noindent
{\it Salla I. Virtanen}
prepared Lipid17ecc parameters.
%% Please write description of your contribution here
